---
title: "Lab2b_CommunityOrdination"
author: "Steven Cognac"
date: "1/24/2022"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

# Learning Objectives
In this lab, you will play with unsupervised classification techniques while working with ecological community datasets.
  
 - **Ordination** orders sites near each other based on similarity. It is a multivariate analysis technique used to effectively collapse dependent axes into fewer dimensions, i.e. dimensionality reduction.
    - **Principal Components Analyses (PCA)** is the most common and oldest technique that assumes linear relationships between axes. You will follow a non-ecological example from [Chapter 17 Principal Components Analysis | Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/pca.html) to learn about this commonly used technique.
    - **Non-metric MultiDimensional Scaling (NMDS)** allows for non-linear relationships. This ordination technique is implemented in the R package [vegan](https://cran.r-project.org/web/packages/vegan/index.html). You’ll use an ecological dataset, species and environment from lichen pastures that reindeer forage upon, with excerpts from the vegantutor vignette (source) to apply these techniques:
      - **Unconstrained ordination** on species using NMDS;
      - Overlay with environmental gradients; and
      - **Constrained ordination** on species and environment using another ordination technique, canonical correspondence analysis (CCA)
  
  
# Ordination
## 1.1 Principal Components Analysis (PCA)
### 1.1.1 Prerequisite
 -  - Prerequisite is to read Chapter 17 (linked above) in entirety.
```{r}
# load R packages
librarian::shelf(
  dplyr, ggplot2, h2o)

# set seed for reproducible results
set.seed(42)

# get data
url <- "https://koalaverse.github.io/homlr/data/my_basket.csv"
my_basket <- readr::read_csv(url)
dim(my_basket)

my_basket

```





